New cardinality estimation algorithms for HyperLogLog sketches

### 2016-05-29 paper/paper.tex
fixed algorithm in paper
use cleveref package, revised chapter 3

\documentclass[a4paper]{scrartcl}
%\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{svg}
\usepackage{hyperref}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

\renewcommand{\sffamily}{\rmfamily}

% ALGORITHMS
\newcommand*{\algoref}[1]{\hyperref[#1]{Algorithm~\ref*{#1}}}

@@

\section{HyperLogLog data structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers is chosen to be a power of 2: $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision that directly influences the relative error scaling like $1/\sqrt{\symNumReg}$. The number of bits per register defines the value range that can be represented. Initially all registers are set equal to 0.

\subsection{Data element insertion}
In order to insert a data element into a HyperLogLog data structure a hash value is calculated. The first $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers. For another $\symRegRange$ bits, the position of the first 1-bit is determined. If this position exceeds the value of the selected register, the register value is replaced. \algoref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.

@@

Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear probabilistic counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm, where $\symRegRange = 0$.

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) initialized to 0 that can take all integers in $[0, \symRegRange+1]$.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem, $\vec{\symRegVal}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\State $\symRegVal' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\State $\symRegVal_\symIndexJ \gets\max(\symRegVal_\symIndexJ, \symRegVal')$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}
Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.

\subsection{Poisson approximation}
Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent. Therefore, as proposed in \cite{Flajolet2007} a Poisson model can be used, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the register values are distributed according to
\begin{align}
\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\nonumber\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexJ=1}^\symNumReg
\symRegProbability_
{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
\nonumber\\
&= 
\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left[
\sum_{\symCardinality=0}^\infty
\symRegProbability_
{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right]^{\symCount_\symRegVal}
\nonumber\\
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\cdot
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
\cdot
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}
\end{align}
This factorization shows that all register values are now independent and identically distributed under the Poisson model. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange
\end{cases}
\end{equation}
Due to the simpler probability mass function, it is easier to find an estimate  $\symPoissonRateEstimate$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Since, the expectation of the cardinality under the Poisson model is equal to $\symPoissonRate$, it makes sense to take $\symPoissonRateEstimate$ direclty as estimate for the cardinality, hence, $\symCardinalityEstimate := \symPoissonRateEstimate$. The Poisson approximation is known to work well for large rates. However, as simulation results show later, the Poisson approximation gives high quality estimates for small cardinalities as well.

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. It is used for the \emph{raw} estimate of the HyperLogLog algorithm and is given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}
\end{equation}
%=
%\left(
%\symNumReg
%\log(2)
%\int_0^1
%v^\symNumReg
%\frac{2^v}{\left(2^v-1\right)^2}
%dv
%\right)^{-1}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision\geq 20$. For HyperLogLog sketches that are used in practice with more than 256 registers ($\symPrecision\geq 8$), it is sufficient to use the limit $\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg$ as approximation 
\begin{equation}
\symAlpha_\symNumReg \approx \symAlpha_\infty = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
because the corresponding approximation error is already small compared to the estimation error.

\autoref{fig:raw_estimate} shows the distribution of the relative error of the \emph{raw} estimate for 10\,000 randomly generated HyperLogLog sketches over the number of distinct element insertions. Obviously, the \emph{raw} estimate is biased for small and large cardinalities and fails to return accurate estimates. As a consequence, to cover the entire range of cardinalities, a small and a large correction have been proposed, respectively.

For small cardinalities the same estimator as used by the linear probabilistic counting algorithm \cite{Whang1990} which only incorporates the number of registers equal to 0 denoted by $\symCount_0$:
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
\frac{\symNumReg}{\symCount_0}
\right)
\end{equation}
The corresponding relative estimation as depicted in \autoref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities.
It was proposed to use the small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
\end{equation}

The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment \emph{raw} estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \autoref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. The empirically determined bias correction data can be either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality as obtained by the raw estimation method \cite{Flajolet2007}. 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the small range correction.}
\label{fig:small_range_estimate}
\end{figure}

@@

see \autoref{fig:raw_corrected_estimation_error_8_24}, \autoref{fig:raw_corrected_estimation_error_12_20},
\autoref{fig:raw_corrected_estimation_error_16_16},
\autoref{fig:raw_corrected_estimation_error_20_12},
\autoref{fig:raw_corrected_estimation_error_22_42},

@@

\subsection{Performance}
see \autoref{fig:corrected_raw_avg_exec_time}

@@

All these optimizations together finally give the new cardinality estimation algorithm presented in \algoref{alg:estimate_ml}.

@@

TODO \autoref{fig:estimation_error_22_42}

>>>>

\documentclass[a4paper]{scrartcl}
%\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{svg}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...
\usepackage{hyperref}
\usepackage[capitalise,nameinlink,noabbrev]{cleveref}

% \renewcommand{\sffamily}{\rmfamily}

@@

\section{HyperLogLog data structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers is chosen to be a power of 2: $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision that directly influences the relative error which scales like $1/\sqrt{\symNumReg}$. All registers start with the initial value 0. The HyperLogLog algorithm successively increases the register values up to a maximum value limited by the space per register which is usually between 5 and 8 bits.

\subsection{Data element insertion}
\label{sec:data_element_insertion}
In order to insert a data element into a HyperLogLog data structure a hash value is calculated. The first $\symPrecision$ bits of the hash value are used to select one of the $2^\symPrecision$ registers. For another $\symRegRange$ bits, the position of the first 1-bit is determined. If this position exceeds the value of the selected register, the register value is replaced. \cref{alg:insert} shows the update procedure for inserting a data element into the HyperLogLog sketch.

@@

Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear probabilistic counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).

\begin{algorithm}
\caption{Insertion of a data element $\symDataItem$ into a HyperLogLog data structure that consists of $\symNumReg=2^\symPrecision$ registers $\vec{\symRegVal} = (\symRegVal_1,\ldots,\symRegVal_\symNumReg$) which are initialized to 0 and which can represent any integer in range $[0, \symRegRange+1]$.}
\label{alg:insert}
\begin{algorithmic}
\Procedure {InsertElement}{\symDataItem, $\vec{\symRegVal}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision,\symBitRepB_1,\ldots,\symBitRepB_\symRegRange\rangle_2 \gets$ $(\symPrecision + \symRegRange)$-bit hash value of $\symDataItem$
\State $\symIndexJ \gets 1+ \langle \symBitRepA_1, \ldots, \symBitRepA_\symPrecision\rangle_2$
\State $\symRegVal' = \min(\{\symS\mid \symS \in [1, \symRegRange]  \wedge  \symBitRepB_\symS = 1\}\cup {\{\symRegRange+1\}} )$
\State $\symRegVal_\symIndexJ \gets\max(\symRegVal_\symIndexJ, \symRegVal')$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}
Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$ and $\symRegRange=\dim(\vec{\symCount})-2$.

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Unfortunately, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult to analyze, because the register values are statistically dependent. Therefore, as proposed in \cite{Flajolet2007} a Poisson model can be used, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\vec{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\vec{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\nonumber\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexJ=1}^\symNumReg
\symRegProbability_
{\symCardinality_\symIndexJ\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexJ}}{\symCardinality_\symIndexJ!\symNumReg^{\symCardinality_\symIndexJ}}
\nonumber\\
&= 
\prod_{\symIndexJ=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability_
{\symCardinality\symRegVal_\symIndexJ}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left[
\sum_{\symCardinality=0}^\infty
\symRegProbability_
{\symCardinality\symRegVal}e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right]^{\symCount_\symRegVal}
\nonumber\\
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\cdot
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\symCount_\symRegVal}\right)
\cdot
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\symCount_{\symRegRange+1}}.
\end{align}
The final factorization shows that all register values are independent and identically distributed under the Poisson model. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is given by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange
\end{cases}
\end{equation}
Due to the simpler probability mass function, it is easier to find an estimate  $\symPoissonRateEstimate$ for the Poisson rate $\symPoissonRate$ than for the cardinality $\symCardinality$ in the fixed-size model \eqref{equ:multinomialProbabilityMass}. Since, the expectation of the cardinality under the Poisson model is equal to $\symPoissonRate$, it makes sense to use $\symPoissonRateEstimate$ directly as estimate for the cardinality $\symCardinalityEstimate := \symPoissonRateEstimate$. The Poisson approximation is known to work well for large rates. However, as simulation results show later, the Poisson approximation gives high quality estimates for small cardinalities as well.

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw HyperLogLog estimate and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision\geq 20$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible when compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimate as function of the cardinality. The chart is based on 10\,000 randomly generated HyperLogLog sketches. Obviously, the raw estimate is biased for small and large cardinalities and fails to return accurate estimates. Therefore, to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed, respectively.

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear probabilistic counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, we can apply the corresponding cardinality estimator \cite{Whang1990}
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
\frac{\symNumReg}{\symCount_0}
\right).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the raw estimate underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
\end{equation}

The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment raw estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. The empirically determined bias correction data can be either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality as for the raw estimator \cite{Flajolet2007}. 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$ have been evaluated.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear probabilistic counting estimator which can be used for small cardinalities.}
\label{fig:small_range_estimate}
\end{figure}

@@

see \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_12_20},
\cref{fig:raw_corrected_estimation_error_16_16},
\cref{fig:raw_corrected_estimation_error_20_12},
\cref{fig:raw_corrected_estimation_error_22_42},

@@

\subsection{Performance}
see \cref{fig:corrected_raw_avg_exec_time}

@@

All these optimizations together finally give the new cardinality estimation algorithm presented in \cref{alg:estimate_ml}.

@@

TODO \cref{fig:estimation_error_22_42}
